Is structure based drug design ready for selectivity optimization?

### 2018-11-28 manuscript.tex
checking in changes to introduction

%
%
%  INTRODUCITON (want to see if this works) 
%
%
Free energy methods have proven useful in aiding structure-based drug design by driving the optimization or maintenance of potency in lead optimization. Alchemical free energy calculations allow for prediction of ligand binding free energies, including all enthalpic and entropic contributions~\citep{Chodera2011-jn}. Advances in atomistic molecular mechanics forcefields~\citep{Huang:J.Comput.Chem.:2013,Maier:J.Chem.TheoryComput.:2015,Harder:J.Chem.TheoryComput.:2016,Cournia:2017ip} have allowed free energy methods to reach a level of accuracy  sufficient for predicting ligand potencies.  In well-behaved kinase systems, free energy calculations have previously been shown to achieve mean unsigned errors of less than 1.0 kcal/mol~\citep{Wang:J.Am.Chem.Soc.:2015,Abel:Curr.Opin.Struct.Biol.:2017,abel2017accelerating} when predicting single target potencies. Beyond kinases, alchemical free energy calculations achieved a mean unsigned error of 0.6 kcal/mol in predicting BRD4 inhibitor potencies~\citep{Aldeghi2015-lf}.

@@

\paragraph{Use of physical modeling to predict selectivity is relatively unexplored}
While predicting selectivity is important for drug discovery, but the utility of free energy methods for predicting this property is relatively unexplored. If there is fortuitous cancellation of errors for closely related systems, free energy methods may be much more accurate than expected given the errors made in predicting the potency for each individual target. Previous work predicting the selectivity of three bromodomain inhibitors accross the bromodomain family achieved promising accuracy for single target potencies of roughly 1 kcal/mol, but does not explicitly evaluate any selectivity metrics~\citep{Aldeghi2017-ox} or look at correlation in the errors made for each bromodomain. For kinases, the selectivity of imatinib for Abl kinase over Src~\citep{Lin2013-ft,Lin2014-iv} and within a family of non-receptor tyrosine kinases~\citep{Lin2013-mu} has been studied extensively using molecular dynamics and free energy calculations. This work focuses on understanding the role reorganization energy plays in the exquisite selectivity of imatinib for Abl over Src despite high similarity between cocrystallized binding mode and kinase conformations, and does not touch on the evaluation of the accuracy of these methods, or their applicationto drug discovery on congeneric series of ligands.

\paragraph{Kinases are an interesting and particularly challenging model system for selectivity predictions}
Kinases are a useful model system to work with for assesing the utility of free energy calculations to predict selectivity. With the approval of imatinib for the treatment of chronic myelogenous leukemia in 2001, targeted small molecule kinase inhibitors (SMKIs) have become a major class of therapeutics in treating cancer and other diseases. 
Currently, there are 43 FDA-approved SMKIs~\citep{fda-approved-kinase-inhibitors}, and it is estimated that kinase targeted therapies account for as much as 50\% of current drug development~\citep{Santos:Nat.Rev.DrugDiscov.:2016}, with many more compounds currently in clinical trials. 
While there have been a number of successes, the current stable of FDA-approved kinase inhibitors targets only a small number of kinases implicated in disease, and the design of new selective kinase inhibitors remains a significant challenge.
Achieving desired selectivity profiles is particularly difficult for kinase targets, making them a system where physical modelling has the potential for a large impact. Achieving selective inhibition of kinases is challenging as there are more than 518 protein kinases~\citep{Volkamer2015-jx,Manning2002-cw} with a highly conserved ATP binding site that is targeted by the majority of SMKIs~\citep{Wu2015-oq}.
While kinase inhibitors have been designed to target kinase-specific subpockets and binding modes to achieve selectivity~\citep{Cowan-Jacob2007-rn,Seeliger2007-jn,Huse2002-ml,Harrison2003-ct,Volkamer2016-sj,ChristmannFranck:2016gka}, previous work has shown that both Type I (binding to the active, DFG-in conformation) and Type II (binding to the inactive, DFG-out conformation) inhibitors display a wide variety of selectivities~\citep{Anastassiadis2011-sm,Davis:Nat.Biotechnol.:2011}, often exhibiting significant binding a number of other targets in addition to their primary target. 
Even FDA-approved inhibitors---often the result of extensive drug development programs---bind to a large number of off-target kinases~\citep{Klaeger2017-jr}.

\paragraph{Assessing the ability of alchemical free energy methods to predict selectivity}
Due to the narrow dynamic range of selectivities and larger experimental errors from the propagation of errors for multiple targets, we anticipate difficulty in predicting selectivity if the errors in the alchemical free energy calculations for two targets are largely uncorrelated, or even anticorrelated. However, correlation in the errors of the calculated free energies on the two targets could lead to a fortuitous cancellation of errors in predicting the selectivity between targets, enabling more accurate than expected calculations. 
Here, we investigate the utility of alchemical free energy calculations for the prediction of selectivity, hereafter taken to mean the $\Delta \Delta$G in binding free energies of the same compound for two targets. We employed state of the art relative free energy calculations~\citep{Wang:J.Am.Chem.Soc.:2015,Abel:2017jt} to predict the selectivities of two different congeneric ligand series~\citep{Shao2013-oe, Blake2016-su}, as well as present a simple numerical model to quantify the expected speed up in selectivity optimization expected for different combinations of per target errors and correlation coefficient values. To tease out the effects of a limited number of experimental measurements, we develop a new Bayesian approach to quantify the uncertainty in the correlation coefficient in the predicted change in selectivity on ligand modification, incorporating all sources of uncertainty and correlation in the computation to separate statistical from force field error. We find that in the closely related systems of CDK2 and CDK9, a high correlation of force field errors suggests that free energy methods can have a significant impact on speeding up selectivity optimization. In the more distantly related case (CDK2/ERK2), limited correlation hampers the ability for free energy methods to speed up selectivity optimization.

@@

\paragraph{Charge Change Free Energy Calculations} 
For ligands where a protonation state change was expected to be relevant to binding based on a small state penalty, Jaguar pKa prediction calculations~\citep{Bochevarov2013-bn} were run to identify protonation state changes with pKas within 1 log unit of the experimental pH. The predicted pKas for one ligand (Shao 12b, 7.84) was within this range. To account for this, a pKa correction was performed. For this ligand, a separate perturbation map containing ligands 12a, 12c, 12b (neutral) and 12b (charged) was run for 30ns per replica using a post-calculation Coulombic charge correction. \textbf{Add in SALT concentration used for these}. The pKa correction was performed using Equation~\ref{eq1}:

>>>>

%
%
%  INTRODUCITON (want to see if this works) 
%
%
Free energy methods have proven useful in aiding structure-based drug design by driving the optimization or maintenance of potency in lead optimization. Alchemical free energy calculations allow for prediction of ligand binding free energies, including all enthalpic and entropic contributions~\citep{Chodera2011-jn}. Advances in atomistic molecular mechanics forcefields and free energy methodologies~\citep{Huang:J.Comput.Chem.:2013,Maier:J.Chem.TheoryComput.:2015,Harder:J.Chem.TheoryComput.:2016,Cournia:2017ip} have allowed free energy methods to reach a level of accuracy  sufficient for predicting ligand potencies~\citep{BROWN2009420}. Free energy methods have been applied prospectively to develop inhibitors for Tyk2~\citep{Abel2017-gw}, Syk~\citep{Lovering:2016fg}, BACE1~\citep{Ciordia:2016dn}, GPCRs~\citep{Lenselink:2016ip}, and HIV protease~\citep{Jorgensen:2016dv}. A recent large-scale review found that the use of FEP+~\citep{Wang:J.Am.Chem.Soc.:2015} to predicting potency for 92 different projects and 3021 compounds found a median RMSE of 1 kcal/mol~\citep{Abel:2017jt}.

@@

\paragraph{Use of physical modeling to predict selectivity is relatively unexplored}
While predicting selectivity is important for drug discovery, but the utility of free energy methods for predicting this property is relatively unexplored. If there is fortuitous cancellation of errors for closely related systems, free energy methods may be much more accurate than expected given the errors made in predicting the potency for each individual target. The selectivity of Imatinib for Abl kinase over Src~\citep{Lin2013-ft,Lin2014-iv} and within a family of non-receptor tyrosine kinases~\citep{Lin2013-mu} has been studied extensively using molecular dynamics and free energy calculations. This work focuses on understanding the role reorganization energy plays in the exquisite selectivity of imatinib for Abl over Src despite high similarity between cocrystallized binding mode and kinase conformations, and does not touch on the evaluation of the accuracy of these methods, or their application to drug discovery on congeneric series of ligands. Previous work predicting the selectivity of three bromodomain inhibitors accross the bromodomain family achieved promising accuracy for single target potencies of roughly 1 kcal/mol, but does not explicitly evaluate any selectivity metrics~\citep{Aldeghi2017-ox} or look at correlation in the errors made for each bromodomain.

\paragraph{Kinases are an interesting and particularly challenging model system for selectivity predictions}
Kinases are a useful model system to work with for assesing the utility of free energy calculations to predict selectivity. With the approval of imatinib for the treatment of chronic myelogenous leukemia in 2001, targeted small molecule kinase inhibitors (SMKIs) have become a major class of therapeutics in treating cancer and other diseases. 
Currently, there are 43 FDA-approved SMKIs~\citep{fda-approved-kinase-inhibitors}, and it is estimated that kinase targeted therapies account for as much as 50\% of current drug development~\citep{Santos:Nat.Rev.DrugDiscov.:2016}, with many more compounds currently in clinical trials. 
While there have been a number of successes, the current stable of FDA-approved kinase inhibitors targets only a small number of kinases implicated in disease, and the design of new selective kinase inhibitors remains a significant challenge.
Achieving desired selectivity profiles is particularly difficult for kinase targets, making them a system where physical modelling has the potential for a large impact. Achieving selective inhibition of kinases is challenging as there are more than 518 protein kinases~\citep{Volkamer2015-jx,Manning2002-cw} with a highly conserved ATP binding site that is targeted by the majority of SMKIs~\citep{Wu2015-oq}.
While kinase inhibitors have been designed to target kinase-specific subpockets and binding modes to achieve selectivity~\citep{Cowan-Jacob2007-rn,Seeliger2007-jn,Huse2002-ml,Harrison2003-ct,Volkamer2016-sj,ChristmannFranck:2016gka}, previous work has shown that both Type I (binding to the active, DFG-in conformation) and Type II (binding to the inactive, DFG-out conformation) inhibitors display a wide variety of selectivities~\citep{Anastassiadis2011-sm,Davis:Nat.Biotechnol.:2011}, often exhibiting significant binding a number of other targets in addition to their primary target. 
Even FDA-approved inhibitors---often the result of extensive drug development programs---bind to a large number of off-target kinases~\citep{Klaeger2017-jr}. Kinases are also targets of interest for developing polypharmacological compounds, or inhibitors that are specifically designed to inhibit multiple kinase targets. Resistance to MEK inhibitors in KRAS-mutant lung and colon cancer has been shown to be driven by HER3 upregulation~\citep{Sun:2014kt}, providing rational for dual MEK/ERBB family inhibitors. Similarly, combined MEK and VEGFR1 inhibition has been proposed as a combinatorial approach to treat KRAS-mutant lung cancer~\citep{Manchado:2016fb}. Developing inhibitors with the desired polypharmacology means navigating more complex selectivity profiles. In well-behaved kinase systems, free energy calculations potency predictions have achieved mean unsigned errors of less than 1.0 kcal/mol~\citep{Wang:J.Am.Chem.Soc.:2015,Abel2017-gw}, suggesting that kinases can be computationally tractable as well as clinically interesting.

\paragraph{Assessing the ability of alchemical free energy methods to predict selectivity}
We anticipate difficulty in predicting selectivity if the errors in the alchemical free energy calculations for two targets are largely uncorrelated, or even anticorrelated. However, correlation in the forcefield errors of the free energies for the two targets could lead to a fortuitous cancellation of errors in predicting the selectivity between targets, making selectivity predictions \emph{more} accurate than potency predictions. Such correlation could occur because the same chemical elements appear in the ligand and in highly related binding sites. 
Here, we investigate the magnitude of this correlation ($\rho$) and the utility of alchemical free energy calculations for the prediction of selectivity, hereafter taken to mean the $\Delta \Delta$G in binding free energies of the same compound for two targets. We employed state of the art relative free energy calculations~\citep{Wang:J.Am.Chem.Soc.:2015,Abel:2017jt} to predict the selectivities of two different congeneric ligand series~\citep{Shao2013-oe, Blake2016-su}, as well as present a simple numerical model to quantify the potential speed up in selectivity optimization expected for different combinations of per target errors and correlation coefficient values. To tease out the effects of a limited number of experimental measurements, we develop a new Bayesian approach to quantify the uncertainty in the correlation coefficient in the predicted change in selectivity on ligand modification, incorporating all sources of uncertainty and correlation in the computation to separate statistical from force field error. We find that in the closely related systems of CDK2 and CDK9, a high correlation of force field errors suggests that free energy methods can have a significant impact on speeding up selectivity optimization. In the more distantly related case (CDK2/ERK2), limited correlation hampers the ability for free energy methods to speed up selectivity optimization.

@@

\paragraph{Charge Change Free Energy Calculations} 
For ligands where a protonation state change was expected to be relevant to binding based on a small state penalty, Jaguar pKa prediction calculations~\citep{Bochevarov2013-bn} were run to identify protonation state changes with pKas within 1 log unit of the experimental pH. The predicted pKas for one ligand (Shao 12b, 7.84) was within this range. To account for this, a pKa correction was performed. For this ligand, a separate perturbation map containing ligands 12a, 12c, 12b (neutral) and 12b (charged) was run for 30ns per replica using a post-calculation Coulombic charge correction. Each charge change pertubation edge was run with a 150mM NaCl concentration. The pKa correction was performed using Equation~\ref{eq1}:
